---
title: "X-linked MethDep in early embryos, Zhu scWGBS data"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---



```{r, echo = FALSE, Results = 'asis'}
suppressPackageStartupMessages(library("dplyr"))
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
library("tidyverse")
library("ComplexHeatmap")
library("CTexploreR")
library("circlize")
library("SummarizedExperiment")
library("SingleCellExperiment")
library("DESeq2")
```

#

```{r}
legends_param <- list(
  labels_gp = gpar(col = "black", fontsize = 3),
  title_gp = gpar(col = "black", fontsize = 3, fontface = "bold"),
  simple_anno_size = unit(0.1, "cm"),
  row_names_gp = gpar(fontsize = 1),
  annotation_name_side = "left",
  border = FALSE,
  border_gp = gpar(lwd = 0.1),
  grid_width = unit(0.1, "cm"),
  grid_height = unit(0.01, "cm"),
  legend_height = unit(0.1, "cm"))

sex_colors<- c("XX" = "deeppink",
               "XY"= "steelblue",
               "X" = "gray20")


CT_genes_X_met <- all_genes %>% 
  filter(CT_gene_type == "CT_gene") %>% 
  filter(X_linked, regulated_by_methylation) %>% 
  pull(external_gene_name)
```


Data from:
[Single-cell DNA methylome sequencing of human preimplantation embryos. Nat Genet 50, 12–19 (2018). Zhu, P., Guo, H., Ren, Y. et al.](https://doi.org/10.1038/s41588-017-0007-6)

# Selection of X-linked CG MethDep genes initially methylated in oocytes

```{r}
non_missing_values_thr <- 5
met_thr_in_oocytes <- 33 #30
met_thr_in_sperm <- 33 #30
```



Selected X-linked CG MethDep genes initially methylated in oocytes



Selection procedure:

- Selected all X-linked genes with at least `r non_missing_values_thr` methylation values in each cell type (oocytes and sperm cells)

- Selected X-linked genes showing at least `r met_thr_in_oocytes`% methylation in MII oocytes (but less than `r met_thr_in_sperm`% in sperm)

- Kept Sperm cells carrying an X chromosome

```{r}
mean_met <- CTdata::mean_methylation_in_embryo()
mean_met <- mean_met[rowRanges(mean_met)$external_gene_name %in% CT_genes_X_met,]

Y_sperm_cells <- as_tibble(colData(mean_met), rownames = "cell") %>% 
  filter(Genotype.of.the.embryo == "Y") %>% 
  pull(cell)

genes_selected_in_oocytes <- as_tibble(assay(mean_met), rownames = "gene") %>% 
  pivot_longer(names_to = "cell", values_to = "met", -gene) %>% 
  left_join(as_tibble(colData(mean_met), rownames = "cell")) %>% 
  filter(!cell %in% Y_sperm_cells) %>% 
  filter(Genotype.of.the.embryo != "XXX") %>% 
  dplyr::rename(Genotype = Genotype.of.the.embryo) %>% 
  filter(!low_quality) %>% 
  filter(Stage %in% c("MII Oocyte")) %>% 
  filter(!is.na(met)) %>% 
  group_by(gene) %>% 
  dplyr::count() %>% 
  filter(n >= non_missing_values_thr) %>% 
  pull(gene)

genes_selected_in_sperm <- as_tibble(assay(mean_met), rownames = "gene") %>% 
  pivot_longer(names_to = "cell", values_to = "met", -gene) %>% 
  left_join(as_tibble(colData(mean_met), rownames = "cell")) %>% 
  filter(!cell %in% Y_sperm_cells) %>% 
  filter(Genotype.of.the.embryo != "XXX") %>% 
  dplyr::rename(Genotype = Genotype.of.the.embryo) %>% 
  filter(!low_quality) %>% 
  filter(Stage %in% c("Sperm"))%>% 
  filter(!is.na(met)) %>% 
  group_by(gene) %>% 
  dplyr::count() %>% 
  filter(n >= non_missing_values_thr) %>% 
  pull(gene)

genes_selected <- BiocGenerics::intersect(genes_selected_in_oocytes, genes_selected_in_sperm)

bilan <- as_tibble(assay(mean_met), rownames = "gene") %>% 
  filter(gene %in% genes_selected) %>% 
  pivot_longer(names_to = "cell", values_to = "met", -gene) %>% 
  left_join(as_tibble(colData(mean_met), rownames = "cell")) %>% 
  filter(!cell %in% Y_sperm_cells) %>% 
  filter(Genotype.of.the.embryo != "XXX") %>% 
  dplyr::rename(Genotype = Genotype.of.the.embryo) %>% 
  filter(!low_quality) %>% 
  filter(Stage %in% c("MII Oocyte", "Sperm")) %>% 
  group_by(gene, Stage, Genotype) %>% 
  dplyr::summarise(mean_met = mean(met, na.rm = TRUE)) %>% 
  mutate(stage_sex = paste0(Stage, '_', Genotype)) 

```






```{r}
genes_methylated_in_oocytes <- bilan %>% 
  filter(Stage %in% c("MII Oocyte")) %>% 
  arrange(mean_met) %>% 
  filter(mean_met >= met_thr_in_oocytes) %>% 
  pull(gene)

met_in_sperm_genes <- bilan %>% 
  filter(Stage %in% c("Sperm")) %>% 
  arrange(mean_met) %>% 
  filter(mean_met > met_thr_in_sperm) %>% 
  pull(gene)

selected_genes <- genes_methylated_in_oocytes[!genes_methylated_in_oocytes %in% met_in_sperm_genes]

bilan %>% 
  mutate(selected = ifelse(gene %in% selected_genes, "kept", "removed")) %>% 
  filter(Stage %in% c("MII Oocyte")) %>% 
  arrange(mean_met) %>% 
  left_join(all_genes %>% dplyr::rename(gene = external_gene_name) %>% dplyr::select(gene, CT_gene_type)) %>% 
  ggplot(aes(x = gene, y = mean_met, color = selected, label = gene)) +
  geom_point() +
  geom_hline(yintercept = met_thr_in_oocytes) +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom")  +
  geom_text( color = "black", size = 3) 
  
```




```{r}
tested_genes <- as_tibble(assay(mean_met), rownames = "gene") %>% 
  filter(gene %in% selected_genes) %>% 
  pull(gene)
tested_genes
length(tested_genes)

bilan_selected <- as_tibble(assay(mean_met), rownames = "gene") %>% 
  filter(gene %in% tested_genes) %>% 
  pivot_longer(names_to = "cell", values_to = "met", -gene) %>% 
  left_join(as_tibble(colData(mean_met), rownames = "cell")) %>% 
  filter(!cell %in% Y_sperm_cells) %>% 
  filter(Genotype.of.the.embryo != "XXX") %>% 
  dplyr::rename(Genotype = Genotype.of.the.embryo) %>% 
  #filter(!low_quality) %>%  # otherwise no more XX-8-cells 
  filter(!Stage %in% c("GV Oocyte")) %>% 
  group_by(gene, Stage, Genotype) %>% 
  dplyr::summarise(mean_met = mean(met, na.rm = TRUE)) %>% 
  mutate(stage_sex = paste0(Stage, '_', Genotype)) 


levels(bilan_selected$Stage)[10] <- "Post-impl"

fig <- bilan_selected %>% 
  ggplot(aes(x = Stage, y = mean_met, group = Genotype)) +
  geom_boxplot(aes(x = Stage, y = mean_met, group = stage_sex, 
                   color = Genotype, fill = Genotype), 
               outliers = FALSE,alpha = 0.1) +
  scale_color_manual(values = sex_colors) +
  scale_fill_manual(values = sex_colors) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12, face = "bold"),
        axis.title.y = element_text(size = 12, face = "bold"),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linewidth = 0.2)) +
  ylab("Mean promoter methylation (%)") +
  xlab('')

fig

pdf(file = paste0("../figs/Fig8A_CG_met_in_Male_and_Female_embryos.pdf"),
    width = 5, height = 4)
fig
dev.off()

```
## t.tests

### Gametes stage

```{r}
t.test(bilan_selected %>% 
  filter(Genotype == "XX") %>% 
  filter(Stage == "MII Oocyte") %>% 
  pull(mean_met),
  bilan_selected %>% 
  filter(Genotype == "X") %>% 
  filter(Stage == "Sperm") %>% 
  pull(mean_met))
```

### at zygote stage

```{r}
stage <- "Zygote"
t.test(bilan_selected %>% 
  filter(Genotype == "XX") %>% 
  filter(Stage == stage) %>% 
  pull(mean_met),
  bilan_selected %>% 
  filter(Genotype == "XY") %>% 
  filter(Stage == stage) %>% 
  pull(mean_met))
```



### at 2-cell stage

```{r}
stage <- "2-cell"
t.test(bilan_selected %>% 
  filter(Genotype == "XX") %>% 
  filter(Stage == stage) %>% 
  pull(mean_met),
  bilan_selected %>% 
  filter(Genotype == "XY") %>% 
  filter(Stage == stage) %>% 
  pull(mean_met))
```



### at 4-cell stage

```{r}
stage <- "4-cell"
t.test(bilan_selected %>% 
  filter(Genotype == "XX") %>% 
  filter(Stage == stage) %>% 
  pull(mean_met),
  bilan_selected %>% 
  filter(Genotype == "XY") %>% 
  filter(Stage == stage) %>% 
  pull(mean_met))
```

### at 8-cell stage

```{r}
stage <- "8-cell"
t.test(bilan_selected %>% 
  filter(Genotype == "XX") %>% 
  filter(Stage == stage) %>% 
  pull(mean_met),
  bilan_selected %>% 
  filter(Genotype == "XY") %>% 
  filter(Stage == stage) %>% 
  pull(mean_met))
```
### at Morula stage

```{r}
stage <- "Morula"
t.test(bilan_selected %>% 
  filter(Genotype == "XX") %>% 
  filter(Stage == stage) %>% 
  pull(mean_met),
  bilan_selected %>% 
  filter(Genotype == "XY") %>% 
  filter(Stage == stage) %>% 
  pull(mean_met))
```


### at Blastocyst stage

```{r}
stage <- "Blastocyst"
t.test(bilan_selected %>% 
  filter(Genotype == "XX") %>% 
  filter(Stage == stage) %>% 
  pull(mean_met),
  bilan_selected %>% 
  filter(Genotype == "XY") %>% 
  filter(Stage == stage) %>% 
  pull(mean_met))
```

### at Post-implantation stage

```{r}
stage <- "Post-impl"
t.test(bilan_selected %>% 
  filter(Genotype == "XX") %>% 
  filter(Stage == stage) %>% 
  pull(mean_met),
  bilan_selected %>% 
  filter(Genotype == "XY") %>% 
  filter(Stage == stage) %>% 
  pull(mean_met))
```

# Example of genes showing a transient imprint in early embryos

```{r}
gamete_colors <- c("MII oocytes" = "deeppink",
               "Sperm"= "steelblue")

MethDep_colors <- c("MethDep" = "indianred2",
"Non-MethDep"= "cyan4")

Stage_col <- c("GV Oocyte" = "mediumvioletred", "MII Oocyte" = "brown4", 
               "Sperm" = "palegreen2", "Zygote" = "lightgoldenrod1", 
               "2-cell" = "ivory2", "4-cell" = "lightslateblue", 
               "8-cell" = "plum2", "Morula" = "deepskyblue",
               "Blastocyst" = "sienna1",  "Post-implantation" = "peachpuff")

stage_level_col <- 
  c("GV Oocyte" = "mediumvioletred", "MII Oocyte" = "brown4", 
    "Sperm" = "palegreen2", 
    "PN-early" = "yellow", "PN-mid" = "gold", "PN-late" = "gold4",
    "2-cell" = "ivory2", "4-cell" = "lightslateblue", 
    "8-cell" = "plum2", "Morula" = "deepskyblue",
    "Blastocyst (Mixed)" = "sienna1",  "Blastocyst (ICM)" = "indianred1",
    "Blastocyst (TE)" = "purple", "Heart" = "steelblue", "Villus" = "gray" )
```


```{r}
methylation_in_embryo <- CTdata::methylation_in_embryo()
load("/storage/research/dduv/cbio-lg/cluster/Packages/CTdata/inst/extdata/all_promoter_regions_hg19.rda")

legends_param <- list(
  labels_gp = gpar(col = "black", fontsize = 10),
  title_gp = gpar(col = "black", fontsize = 10, fontface = "bold"),
  simple_anno_size = unit(0.1, "cm"),
  row_names_gp = gpar(fontsize = 1),
  annotation_name_side = "left",
  border = FALSE,
  border_gp = gpar(lwd = 0.1),
  grid_width = unit(0.3, "cm"),
  grid_height = unit(0.3, "cm"),
  legend_height = unit(0.5, "cm"))
```


```{r}

promoter_methylation_by_GR <- function(GR = NULL, 
                                       nt_up = 1000,
                                       nt_down = 500,
                                       sc_or_bulk = NULL, # NULL, single or bulk
                                       genotype = NULL,
                                       stage = NULL,
                                       split_by = "stage_level",
                                       rm_low_quality = TRUE,
                                       rm_cell_with_only_missing_values = TRUE,
                                       clustRow = FALSE,
                                       fs = 6){

  x <- GR
  gene <- x$external_gene_name
  tested_gene_GR <- subsetByOverlaps(methylation_in_embryo, x)
  CpGpos <- tested_gene_GR@rowRanges@ranges@start
  TSS_hg19 <- x$TSS_liftover
  if(as.vector(x@strand) == '+'){
    CpGpositions <- CpGpos - x$TSS_liftover
  }
  if(as.vector(x@strand) == '-'){
    CpGpositions <- x$TSS_liftover  - CpGpos 
  }
  df_col <- colData(tested_gene_GR)
  if (rm_low_quality) df_col <- df_col[!is.na(df_col$low_quality) & !df_col$low_quality,]
  if (!is.null(stage)) df_col <- df_col[df_col$stage_level %in% stage, ]
  if (!is.null(genotype)) df_col <- df_col[df_col$`Genotype of the embryo` %in% genotype, ]
  df_col <- df_col[!df_col$`Genotype of the embryo` %in% "XXX", ]
  if (!is.null(sc_or_bulk)){
    if(sc_or_bulk == "single"){
      df_col <- df_col[df_col$`Single Cell or Bulk cell` == "single cell", ]
    }
    if(sc_or_bulk == "bulk"){
      df_col <- df_col[df_col$`Single Cell or Bulk cell` == "bulk cells", ]
    }
  }
  
  df_col <- df_col[order(df_col$stage_level, df_col$`Genotype of the embryo`),] 
  
  y <- as_tibble(assay(tested_gene_GR)[, df_col$Sample_Name]) %>% 
    mutate(CpG = CpGpositions) %>% dplyr::select(CpG, everything()) %>% 
    filter(CpG > -nt_up & CpG < nt_down) %>% 
   #  filter(CpG > -nt_arround_TS & CpG < nt_arround_TS) %>% 
    arrange(CpG) %>% 
    pivot_longer(names_to = "Sample_Name", values_to = "met", -CpG) %>% 
    pivot_wider(names_from = CpG, values_from = met)
  mat <- as.matrix(y[, -1])
  rownames(mat) <- y$Sample_Name

  if (rm_cell_with_only_missing_values) {
    if (length(which(rowSums(is.na(mat)) == ncol(mat))) > 0) {
      mat <- mat[-which(rowSums(is.na(mat)) == ncol(mat)), ,drop = F]
      df_col <- df_col[rownames(mat),]
    }
  }
  
  rowSplit <- df_col$stage_level
  if (split_by == "Stage") {
    df_col <- df_col[order(df_col$Stage, df_col$`Genotype of the embryo`),] 
    rowSplit <- df_col$Stage
  }
  
  if (split_by == "embryo") {
    df_col$embryo <- paste0(df_col$stage_level, "_", df_col$individual)
    df_col$embryo[df_col$stage_level %in% c("GV Oocyte")] <- "GV Oocyte"
    df_col$embryo[df_col$stage_level %in% c("MII Oocyte")] <- "MII Oocyte"
    df_col$embryo[df_col$stage_level %in% c("Sperm")] <- "Sperm"
    df_col$embryo[df_col$stage_level %in% c("2-cell")] <- "2-cell"
        df_col$embryo[df_col$stage_level %in% c("4-cell")] <- "4-cell"
            df_col$embryo[df_col$stage_level %in% c("8-cell")] <- "8-cell"
    df_col$embryo[df_col$stage_level %in% c("Heart")] <- "Heart"
    df_col$embryo[df_col$stage_level %in% c("Villus")] <- "Villus"
    df_col <- df_col[order(df_col$stage_level, df_col$`Genotype of the embryo`, df_col$embryo),] 
    rowSplit <- factor(df_col$embryo, levels = unique(df_col$embryo))
  }

  Genotype <- as.matrix(df_col$`Genotype of the embryo`)
  rownames(Genotype) <- df_col$Sample_Name
  row_ha_Genotype <- rowAnnotation(Genotype = Genotype,
                              na_col = "gray80",
                              annotation_legend_param = legends_param,
                              simple_anno_size = unit(0.2, "cm"),
                              col = list(Genotype = c("XX" = "deeppink",
                                                 "XY"= "steelblue",
                                                 "X" = "pink",
                                                 "Y"= "dodgerblue",
                                                 "XXX" = "orange")),
                              annotation_name_gp = gpar(fontsize = 0))

  Stage <- as.matrix(df_col$Stage)
  rownames(Stage) <- df_col$Stage
  row_ha_Stage <- rowAnnotation(Stage = Stage,
                              na_col = "gray80",
                              annotation_legend_param = legends_param,
                              simple_anno_size = unit(0.2, "cm"),
                             # col = list(Stage = Stage_col),
                              annotation_name_gp = gpar(fontsize = 0))
  
  stage_level <- as.matrix(df_col$stage_level)
  rownames(stage_level) <- df_col$stage_level
  row_ha_stage_level <- rowAnnotation(stage_level = stage_level,
                              na_col = "gray80",
                              annotation_legend_param = legends_param,
                              simple_anno_size = unit(0.2, "cm"),
                              col = list(stage_level = stage_level_col),
                              annotation_name_gp = gpar(fontsize = 0))
  data_type <- as.matrix(df_col$bulk_or_single_cell)
  rownames(data_type) <- df_col$bulk_or_single_cell
  row_ha_data_type <- rowAnnotation(data_type = data_type,
                              na_col = "gray80",
                              annotation_legend_param = legends_param,
                              simple_anno_size = unit(0.2, "cm"),
                              col = list(data_type = c("bulk cells" = "black",
                                         "single cell" = "cyan")),
                              annotation_name_gp = gpar(fontsize = 0))
  

  
  h <- Heatmap(mat[rownames(df_col), ],
               column_title = paste0('Promoter methylation of ', gene, " TSS ", TSS_hg19, " (hg19)"),
               name = 'Met',
               row_split = rowSplit,
               cluster_row_slices = FALSE,
               row_gap = unit(c(1), "mm"),
               row_title_gp = gpar(fontsize = 8, fontface = "bold"),
               row_title_rot = 0,
               col = colorRamp2(c(1:100),colorRampPalette(
                 c("moccasin","dodgerblue4"))(100)),
               na_col = "gray80",
               clustering_method_rows = "ward.D",
               cluster_rows = clustRow,
               cluster_columns = FALSE,
               show_row_names = TRUE,
               show_column_names = TRUE,
               show_heatmap_legend = TRUE,
               show_row_dend = FALSE,
               right_annotation = c(row_ha_Genotype,  #row_ha_Stage,
                                    #row_ha_stage_level,
                                    #row_ha_data_type,
                                      gap = unit(1, "mm")),
               row_names_gp = gpar(fontsize = fs),
               column_names_gp = gpar(fontsize = fs),
               column_names_side = "bottom",
               row_names_side = "left")
  
  print(h) 
}
```


```{r}
promoter_methylation_fig <- function(GR = NULL, 
                                     title = "",
                                     nt_up = 1000,
                                     nt_down = 500,
                                     sc_or_bulk = NULL, # NULL, single or bulk
                                     genotype = NULL,
                                     stage = NULL,
                                     split_by = "stage_level",
                                     rm_low_quality = TRUE,
                                     rm_cell_with_only_missing_values = TRUE,
                                     show_legend = TRUE,
                                     cell_per_type = NULL,
                                     clustRow = FALSE,
                                     h_width = 5,
                                     h_height = 10,
                                     rowtitle_fs = 7,
                                     coltitle_fs = 7,
                                     CpG_n_fs = 4,
                                     fs = 6){
  x <- GR
  gene <- x$external_gene_name
  tested_gene_GR <- subsetByOverlaps(methylation_in_embryo, x)
  CpGpos <- tested_gene_GR@rowRanges@ranges@start
  TSS_hg19 <- x$TSS_liftover
  if(as.vector(x@strand) == '+'){
    CpGpositions <- CpGpos - x$TSS_liftover
  }
  if(as.vector(x@strand) == '-'){
    CpGpositions <- x$TSS_liftover  - CpGpos 
  }
  
  df_col <- colData(tested_gene_GR)
  if (rm_low_quality) df_col <- df_col[!is.na(df_col$low_quality) & !df_col$low_quality,]
  if (!is.null(stage)) df_col <- df_col[df_col$stage_level %in% stage, ]
  if (!is.null(genotype)) df_col <- df_col[df_col$`Genotype of the embryo` %in% genotype, ]
  df_col <- df_col[!df_col$`Genotype of the embryo` %in% "XXX", ]
  
  if (!is.null(sc_or_bulk)){
    if(sc_or_bulk == "single"){
      df_col <- df_col[df_col$`Single Cell or Bulk cell` == "single cell", ]
    }
    if(sc_or_bulk == "bulk"){
      df_col <- df_col[df_col$`Single Cell or Bulk cell` == "bulk cells", ]
    }
  }
  
  df_col <- df_col[!df_col$Sample_Name %in% c("Icm03-W-s64","scBS-C-MII-9"), ]
  
  
  y <- as_tibble(assay(tested_gene_GR)[, df_col$Sample_Name]) %>% 
    mutate(CpG = CpGpositions) %>% dplyr::select(CpG, everything()) %>% 
    filter(CpG > -nt_up & CpG < nt_down) %>% 
    #  filter(CpG > -nt_arround_TS & CpG < nt_arround_TS) %>% 
    arrange(CpG) %>% 
    pivot_longer(names_to = "Sample_Name", values_to = "met", -CpG) %>% 
    pivot_wider(names_from = CpG, values_from = met)
  mat <- as.matrix(y[, -1])
  rownames(mat) <- y$Sample_Name
  
  
  # calculate mean in each sample, to be able to order sequences within in group
  # in fonction of the methylation level
  df_col$median_met <- rowMedians(mat, na.rm = T)
  df_col$prop_of_missing_values <- (rowSums(is.na(mat)) / ncol(mat))
  
  if (rm_cell_with_only_missing_values) df_col <- df_col[df_col$prop_of_missing_values < 1, ]
  
  if (!is.null(cell_per_type)){
    ### Keep n max cell per group, keeping the samples with the less missing values
    
    kept <- as_tibble(df_col) %>% 
      filter(Stage %in% cell_per_type$Stage[1]) %>% 
      arrange(prop_of_missing_values) %>% 
      head(pull(cell_per_type[1, 2])) %>% 
      pull(Sample_Name)
    for (i in seq_along(cell_per_type$Stage)[-1]){
      tmp <- as_tibble(df_col) %>% 
        filter(Stage %in% cell_per_type$Stage[i]) %>% 
        arrange(prop_of_missing_values) %>% 
        head(pull(cell_per_type[i, 2])) %>% 
        pull(Sample_Name)
      kept <- c(kept, tmp)
    }
    df_col <- df_col[df_col$Sample_Name %in% kept, ] 
  }
  
  df_col <- df_col[order(df_col$stage_level, df_col$`Genotype of the embryo`, df_col$median_met),] 
  
  rowSplit <- df_col$stage_level
  
  if (split_by == "Stage") {
    df_col <- df_col[order(df_col$Stage, df_col$`Genotype of the embryo`, df_col$median_met),] 
    rowSplit <- df_col$Stage
  }
  
  
  if (split_by == "embryo") {
    df_col$embryo <- paste0(df_col$stage_level, "_", df_col$individual)
    df_col$embryo[df_col$stage_level %in% c("GV Oocyte")] <- "GV Oocyte"
    df_col$embryo[df_col$stage_level %in% c("MII Oocyte")] <- "MII Oocyte"
    df_col$embryo[df_col$stage_level %in% c("Sperm")] <- "Sperm"
    df_col$embryo[df_col$stage_level %in% c("2-cell")] <- "2-cell"
    df_col$embryo[df_col$stage_level %in% c("4-cell")] <- "4-cell"
    df_col$embryo[df_col$stage_level %in% c("8-cell")] <- "8-cell"
    df_col$embryo[df_col$stage_level %in% c("Heart")] <- "Heart"
    df_col$embryo[df_col$stage_level %in% c("Villus")] <- "Villus"
    df_col <- df_col[order(df_col$stage_level, df_col$`Genotype of the embryo`, df_col$embryo, df_col$mean_met),] 
    rowSplit <- factor(df_col$embryo, levels = unique(df_col$embryo))
  }
  
  Genotype <- as.matrix(df_col$`Genotype of the embryo`)
  rownames(Genotype) <- df_col$Sample_Name
  row_ha_Genotype <- rowAnnotation(Genotype = Genotype,
                                   na_col = "gray80",
                                   annotation_legend_param = legends_param,
                                   simple_anno_size = unit(0.2, "cm"),
                                   col = list(Genotype = c("XX" = "deeppink",
                                                           "XY"= "steelblue",
                                                           "X" = "pink",
                                                           "Y"= "dodgerblue",
                                                           "XXX" = "orange")),
                                   annotation_name_gp = gpar(fontsize = 0))
  
  
  Stage <- as.matrix(df_col$Stage)
  rownames(Stage) <- df_col$Stage
  row_ha_Stage <- rowAnnotation(Stage = Stage,
                                na_col = "gray80",
                                annotation_legend_param = legends_param,
                                simple_anno_size = unit(0.2, "cm"),
                                col = list(Stage = Stage_col),
                                annotation_name_gp = gpar(fontsize = 0))
  
  stage_level <- as.matrix(df_col$stage_level)
  rownames(stage_level) <- df_col$stage_level
  row_ha_stage_level <- rowAnnotation(stage_level = stage_level,
                                       na_col = "gray80",
                                       annotation_legend_param = legends_param,
                                       simple_anno_size = unit(0.2, "cm"),
                                       col = list(stage_level = stage_level_col),
                                       annotation_name_gp = gpar(fontsize = 0))
  data_type <- as.matrix(df_col$bulk_or_single_cell)
  rownames(data_type) <- df_col$bulk_or_single_cell
  row_ha_data_type <- rowAnnotation(data_type = data_type,
                                    na_col = "gray80",
                                    annotation_legend_param = legends_param,
                                    simple_anno_size = unit(0.2, "cm"),
                                    col = list(data_type = c("bulk cells" = "black",
                                                             "single cell" = "cyan")),
                                    annotation_name_gp = gpar(fontsize = 0))
  
  
  if (all(all_of(df_col$`Genotype of the embryo`) %in% c("X", "XX"))) gender <- "Female"
  if (all(all_of(df_col$`Genotype of the embryo`) %in% c("Y", "XY"))) gender <- "Male"
  
  ht_opt$TITLE_PADDING = unit(c(2.5, 5), "points")
  h <- Heatmap(mat[rownames(df_col), ],
               column_title = title,
               name = "Met",
               row_split = rowSplit,
               cluster_row_slices = FALSE,
               row_gap = unit(c(3), "mm"),
               border = TRUE,
               border_gp = gpar(lwd = 0.1),
               row_title_gp = gpar( fontsize = rowtitle_fs, fontface = "bold"),
               column_title_gp = gpar(fontsize = coltitle_fs, fontface = "bold"),
               row_title_rot = 90,
               col = colorRamp2(c(1:100),colorRampPalette(
                 c("moccasin","dodgerblue4"))(100)),
               na_col = "gray80",
               clustering_method_rows = "ward.D",
               cluster_rows = clustRow,
               cluster_columns = FALSE,
               show_row_names = FALSE,
               show_column_names = TRUE,
               show_heatmap_legend = show_legend,
               show_row_dend = FALSE,
               row_names_gp = gpar(fontsize = fs),
               column_names_gp = gpar(fontsize = CpG_n_fs),
               heatmap_width = unit(h_width, "cm"),
               heatmap_height = unit(h_height, "cm"),
               column_names_side = "bottom",
               row_names_side = "left",
               heatmap_legend_param = legends_param)
  
  print(h) 
}


```

# MAGEC2 Figures


```{r}
my_gene <- "MAGEC2"
my_GR <- all_promoter_regions_hg19[all_promoter_regions_hg19$external_gene_name == my_gene,]

promoter_methylation_by_GR(GR = my_GR[1], 
                           nt_up = 1000,
                           nt_down = 500,
                           sc_or_bulk = NULL,
                           #genotype = "XX",
                           #stage = "MII Oocyte",
                           #split_by = "individual",
                           split_by = "Stage",
                           rm_low_quality = FALSE,
                           rm_cell_with_only_missing_values = TRUE,
                           clustRow = FALSE,
                           fs = 2)
```

Show Promoter region -/- 1000 pb around TSS

```{r}

prom_region <- c(1000, 1000)

my_GR <- all_promoter_regions_hg19[all_promoter_regions_hg19$external_gene_name == my_gene,]
gene <- my_GR$external_gene_name

cell_per_type <- tibble(Stage = c("MII Oocyte", "Sperm"),
                        n = c(10, 10))


A <- cowplot::plot_grid(grid.grabExpr(
  draw(promoter_methylation_fig(GR = my_GR[1], 
            title = "Oocytes",
            show_legend = FALSE,
            rowtitle_fs = 0,
            CpG_n_fs = 2.5,
            nt_up = prom_region[1],
            nt_down = prom_region[2],
            sc_or_bulk = NULL,
            genotype = c("X", "XX"),
            stage = c("MII Oocyte") ,
            #split_by = "individual",
            split_by = "Stage",
            rm_low_quality = FALSE,
            rm_cell_with_only_missing_values = TRUE,
            cell_per_type = cell_per_type,
            clustRow = FALSE,
            h_width = 4,
            h_height = 2,
            fs = 0)),  labels = c('A')))

B <- cowplot::plot_grid(grid.grabExpr(
  draw(promoter_methylation_fig(GR = my_GR[1], 
            title = "Sperm",
            show_legend = FALSE,
            rowtitle_fs = 0,
             CpG_n_fs = 2.5,
             nt_up = prom_region[1],
            nt_down = prom_region[2],
            sc_or_bulk = NULL,
            genotype = c("X", "XX"),
            stage = c("Sperm") ,
            #split_by = "individual",
            split_by = "Stage",
            rm_low_quality = FALSE,
            rm_cell_with_only_missing_values = TRUE,
            cell_per_type = cell_per_type,
            clustRow = FALSE,
            h_width = 4,
            h_height = 2,
            fs = 0)),  labels = c('A')))

figA <- cowplot::plot_grid(A, B, 
                   # rel_widths = c(2,1),
                   # scale = c(1, 0.8),
                   # label_size = 10, 
                   nrow = 2,
                   align = "h")

# pdf(file = paste0("../figs/Fig8C_", gene, "_Methylation_Oocytes_Sperm.pdf"))#, height = 3, width =4)
# figA
# dev.off()


# Show the same number of cells in male and female embryos
# Define the number of cells per stage manually
cell_per_type <- tibble(Stage = c("Zygote", #"2-cell", "4-cell", "8-cell", 
                                  "Morula", "Blastocyst", "Post-implantation"),
                        n = c(6, #8, 8, 8, 
                              10, 12, 4))


cell_per_type <- tibble(Stage = c("Zygote", #"2-cell", "4-cell", "8-cell", 
                                  "Morula", "Blastocyst", "Post-implantation"),
                        n = c(10, #8, 8, 8, 
                              10, 20, 4))

D <- cowplot::plot_grid(grid.grabExpr(draw(
  promoter_methylation_fig(GR = my_GR[1], 
                           title = "Female",
                           show_legend = FALSE,
                           rowtitle_fs = 8,
                           coltitle_fs = 10,
                           CpG_n_fs = 2.5,
                           nt_up = prom_region[1],
                           nt_down = prom_region[2],
                           sc_or_bulk = "single",
                           genotype = c("X", "XX"),
                           stage = c("PN-early", "PN-mid", "PN-late", "2-cell", #"4-cell", 
                                     "8-cell", "Morula", "Blastocyst (Mixed)", 
                                     "Blastocyst (ICM)", "Heart", "Villus"),    
                           #split_by = "individual",
                           split_by = "Stage",
                           rm_low_quality = FALSE, # otherwise not enough samples
                           rm_cell_with_only_missing_values = TRUE,
                           cell_per_type = cell_per_type,
                           clustRow = FALSE,
                           h_width = 4,
                           h_height =5,
                           fs = 8)),  labels = c('A')))

E <- cowplot::plot_grid(grid.grabExpr(draw(
  promoter_methylation_fig(GR = my_GR[1], 
                           title = "Male",
                           rowtitle_fs = 8,
                           coltitle_fs = 10,
                          CpG_n_fs = 2.5,
                           nt_up = prom_region[1],
                           nt_down = prom_region[2],
                           sc_or_bulk = "single",
                           genotype = c("Y", "XY"),
                           stage = c("PN-early", "PN-mid", "PN-late", "2-cell", "4-cell", 
                                     "8-cell", 
                                     "Morula", "Blastocyst (Mixed)", "Blastocyst (ICM)", "Heart", "Villus"),       
                           #split_by = "individual",
                           split_by = "Stage",
                           rm_low_quality = FALSE, # otherwise not enough samples
                           rm_cell_with_only_missing_values = TRUE,
                           cell_per_type = cell_per_type,
                           clustRow = FALSE,
                           h_width = 4,
                           h_height =5,
                           fs = 8)),  labels = c('B')))

figB<- cowplot::plot_grid(D,E, 
                   # rel_widths = c(2,1),
                   # scale = c(1, 0.8),
                   # label_size = 10, 
                   nrow = 1,
                   align = "h")

# pdf(file = paste0("../figs/Fig8C_", gene, "_Methylation_embryos.pdf"))#, height = 3, width = 3)
# figB
# dev.off()

figC <- cowplot::plot_grid(figA, figB,
                    rel_heights = c(1,1.6),
                   # scale = c(1, 0.8),
                   # label_size = 10, 
                   nrow = 2,
                   align = "h",
                   labels = my_gene)


pdf(file = paste0("../figs/Fig8C_", gene, "_Methylation_embryos.pdf"), height = 3.5, width = 4.5)
figC
dev.off()

figC
```






# DDX53 Figures


```{r}
my_gene <- "DDX53"
my_GR <- all_promoter_regions_hg19[all_promoter_regions_hg19$external_gene_name == my_gene,]

promoter_methylation_by_GR(GR = my_GR[1], 
                           nt_up = 1000,
                           nt_down = 500,
                           sc_or_bulk = NULL,
                           #genotype = "XX",
                           #stage = "MII Oocyte",
                           #split_by = "individual",
                           split_by = "Stage",
                           rm_low_quality = FALSE,
                           rm_cell_with_only_missing_values = TRUE,
                           clustRow = FALSE,
                           fs = 2)
```

Show Promoter region -/- 1000 pb around TSS

```{r}
prom_region <- c(1000, 1000)

my_GR <- all_promoter_regions_hg19[all_promoter_regions_hg19$external_gene_name == my_gene,]
gene <- my_GR$external_gene_name

cell_per_type <- tibble(Stage = c("MII Oocyte", "Sperm"),
                        n = c(10, 10))


A <- cowplot::plot_grid(grid.grabExpr(
  draw(promoter_methylation_fig(GR = my_GR[1], 
            title = "Oocytes",
            show_legend = FALSE,
            rowtitle_fs = 0,
            CpG_n_fs = 2.5,
            nt_up = prom_region[1],
            nt_down = prom_region[2],
            sc_or_bulk = NULL,
            genotype = c("X", "XX"),
            stage = c("MII Oocyte") ,
            #split_by = "individual",
            split_by = "Stage",
            rm_low_quality = FALSE,
            rm_cell_with_only_missing_values = TRUE,
            cell_per_type = cell_per_type,
            clustRow = FALSE,
            h_width = 4,
            h_height = 2,
            fs = 0)),  labels = c('A')))

B <- cowplot::plot_grid(grid.grabExpr(
  draw(promoter_methylation_fig(GR = my_GR[1], 
            title = "Sperm",
            show_legend = FALSE,
            rowtitle_fs = 0,
             CpG_n_fs = 2.5,
             nt_up = prom_region[1],
            nt_down = prom_region[2],
            sc_or_bulk = NULL,
            genotype = c("X", "XX"),
            stage = c("Sperm") ,
            #split_by = "individual",
            split_by = "Stage",
            rm_low_quality = FALSE,
            rm_cell_with_only_missing_values = TRUE,
            cell_per_type = cell_per_type,
            clustRow = FALSE,
            h_width = 4,
            h_height = 2,
            fs = 0)),  labels = c('A')))

figA <- cowplot::plot_grid(A, B, 
                   # rel_widths = c(2,1),
                   # scale = c(1, 0.8),
                   # label_size = 10, 
                   nrow = 2,
                   align = "h")

# pdf(file = paste0("../figs/Fig8C_", gene, "_Methylation_Oocytes_Sperm.pdf"))#, height = 3, width =4)
# figA
# dev.off()


# Show the same number of cells in male and female embryos
# Define the number of cells per stage manually
cell_per_type <- tibble(Stage = c("Zygote", #"2-cell", "4-cell", "8-cell", 
                                  "Morula", "Blastocyst", "Post-implantation"),
                        n = c(6, #8, 8, 8, 
                              10, 12, 4))


cell_per_type <- tibble(Stage = c("Zygote", #"2-cell", "4-cell", "8-cell", 
                                  "Morula", "Blastocyst", "Post-implantation"),
                        n = c(10, #8, 8, 8, 
                              10, 20, 4))

D <- cowplot::plot_grid(grid.grabExpr(draw(
  promoter_methylation_fig(GR = my_GR[1], 
                           title = "Female",
                           show_legend = FALSE,
                           rowtitle_fs = 8,
                           coltitle_fs = 10,
                           CpG_n_fs = 2.5,
                           nt_up = prom_region[1],
                           nt_down = prom_region[2],
                           sc_or_bulk = "single",
                           genotype = c("X", "XX"),
                           stage = c("PN-early", "PN-mid", "PN-late", "2-cell", #"4-cell", 
                                     "8-cell", "Morula", "Blastocyst (Mixed)", 
                                     "Blastocyst (ICM)", "Heart", "Villus"),    
                           #split_by = "individual",
                           split_by = "Stage",
                           rm_low_quality = FALSE, # otherwise not enough samples
                           rm_cell_with_only_missing_values = TRUE,
                           cell_per_type = cell_per_type,
                           clustRow = FALSE,
                           h_width = 4,
                           h_height =5,
                           fs = 8)),  labels = c('A')))

E <- cowplot::plot_grid(grid.grabExpr(draw(
  promoter_methylation_fig(GR = my_GR[1], 
                           title = "Male",
                           rowtitle_fs = 8,
                           coltitle_fs = 10,
                          CpG_n_fs = 2.5,
                           nt_up = prom_region[1],
                           nt_down = prom_region[2],
                           sc_or_bulk = "single",
                           genotype = c("Y", "XY"),
                           stage = c("PN-early", "PN-mid", "PN-late", "2-cell", "4-cell", 
                                     "8-cell", 
                                     "Morula", "Blastocyst (Mixed)", "Blastocyst (ICM)", "Heart", "Villus"),       
                           #split_by = "individual",
                           split_by = "Stage",
                           rm_low_quality = FALSE, # otherwise not enough samples
                           rm_cell_with_only_missing_values = TRUE,
                           cell_per_type = cell_per_type,
                           clustRow = FALSE,
                           h_width = 4,
                           h_height =5,
                           fs = 8)),  labels = c('B')))

figB<- cowplot::plot_grid(D,E, 
                   # rel_widths = c(2,1),
                   # scale = c(1, 0.8),
                   # label_size = 10, 
                   nrow = 1,
                   align = "h")

# pdf(file = paste0("../figs/Fig8C_", gene, "_Methylation_embryos.pdf"))#, height = 3, width = 3)
# figB
# dev.off()


figC <- cowplot::plot_grid(figA, figB,
                    rel_heights = c(1,1.6),
                   # scale = c(1, 0.8),
                   # label_size = 10, 
                   nrow = 2,
                   align = "h",
                   labels = my_gene)


pdf(file = paste0("../figs/Fig8C_", gene, "_Methylation_embryos.pdf"), height = 3.5, width = 4.5)
figC
dev.off()

figC
```







# MAGEB6 Figures


```{r}
my_gene <- "MAGEB6"
my_GR <- all_promoter_regions_hg19[all_promoter_regions_hg19$external_gene_name == my_gene,]

promoter_methylation_by_GR(GR = my_GR[1], 
                           nt_up = 1000,
                           nt_down = 500,
                           sc_or_bulk = NULL,
                           #genotype = "XX",
                           #stage = "MII Oocyte",
                           #split_by = "individual",
                           split_by = "Stage",
                           rm_low_quality = FALSE,
                           rm_cell_with_only_missing_values = TRUE,
                           clustRow = FALSE,
                           fs = 2)
```

Show Promoter region -/- 1000 pb around TSS

```{r}
prom_region <- c(1000, 1000)

my_GR <- all_promoter_regions_hg19[all_promoter_regions_hg19$external_gene_name == my_gene,]
gene <- my_GR$external_gene_name

cell_per_type <- tibble(Stage = c("MII Oocyte", "Sperm"),
                        n = c(10, 10))


A <- cowplot::plot_grid(grid.grabExpr(
  draw(promoter_methylation_fig(GR = my_GR[1], 
            title = "Oocytes",
            show_legend = FALSE,
            rowtitle_fs = 0,
            CpG_n_fs = 2.5,
            nt_up = prom_region[1],
            nt_down = prom_region[2],
            sc_or_bulk = NULL,
            genotype = c("X", "XX"),
            stage = c("MII Oocyte") ,
            #split_by = "individual",
            split_by = "Stage",
            rm_low_quality = FALSE,
            rm_cell_with_only_missing_values = TRUE,
            cell_per_type = cell_per_type,
            clustRow = FALSE,
            h_width = 4,
            h_height = 2,
            fs = 0)),  labels = c('A')))

B <- cowplot::plot_grid(grid.grabExpr(
  draw(promoter_methylation_fig(GR = my_GR[1], 
            title = "Sperm",
            show_legend = FALSE,
            rowtitle_fs = 0,
             CpG_n_fs = 2.5,
             nt_up = prom_region[1],
            nt_down = prom_region[2],
            sc_or_bulk = NULL,
            genotype = c("X", "XX"),
            stage = c("Sperm") ,
            #split_by = "individual",
            split_by = "Stage",
            rm_low_quality = FALSE,
            rm_cell_with_only_missing_values = TRUE,
            cell_per_type = cell_per_type,
            clustRow = FALSE,
            h_width = 4,
            h_height = 2,
            fs = 0)),  labels = c('A')))

figA <- cowplot::plot_grid(A, B, 
                   # rel_widths = c(2,1),
                   # scale = c(1, 0.8),
                   # label_size = 10, 
                   nrow = 2,
                   align = "h")

# pdf(file = paste0("../figs/Fig8C_", gene, "_Methylation_Oocytes_Sperm.pdf"))#, height = 3, width =4)
# figA
# dev.off()


# Show the same number of cells in male and female embryos
# Define the number of cells per stage manually
cell_per_type <- tibble(Stage = c("Zygote", #"2-cell", "4-cell", "8-cell", 
                                  "Morula", "Blastocyst", "Post-implantation"),
                        n = c(6, #8, 8, 8, 
                              10, 12, 4))


cell_per_type <- tibble(Stage = c("Zygote", #"2-cell", "4-cell", "8-cell", 
                                  "Morula", "Blastocyst", "Post-implantation"),
                        n = c(10, #8, 8, 8, 
                              10, 20, 4))

D <- cowplot::plot_grid(grid.grabExpr(draw(
  promoter_methylation_fig(GR = my_GR[1], 
                           title = "Female",
                           show_legend = FALSE,
                           rowtitle_fs = 8,
                           coltitle_fs = 10,
                           CpG_n_fs = 2.5,
                           nt_up = prom_region[1],
                           nt_down = prom_region[2],
                           sc_or_bulk = "single",
                           genotype = c("X", "XX"),
                           stage = c("PN-early", "PN-mid", "PN-late", "2-cell", #"4-cell", 
                                     "8-cell", "Morula", "Blastocyst (Mixed)", 
                                     "Blastocyst (ICM)", "Heart", "Villus"),    
                           #split_by = "individual",
                           split_by = "Stage",
                           rm_low_quality = FALSE, # otherwise not enough samples
                           rm_cell_with_only_missing_values = TRUE,
                           cell_per_type = cell_per_type,
                           clustRow = FALSE,
                           h_width = 4,
                           h_height =5,
                           fs = 8)),  labels = c('A')))

E <- cowplot::plot_grid(grid.grabExpr(draw(
  promoter_methylation_fig(GR = my_GR[1], 
                           title = "Male",
                           rowtitle_fs = 8,
                           coltitle_fs = 10,
                          CpG_n_fs = 2.5,
                           nt_up = prom_region[1],
                           nt_down = prom_region[2],
                           sc_or_bulk = "single",
                           genotype = c("Y", "XY"),
                           stage = c("PN-early", "PN-mid", "PN-late", "2-cell", "4-cell", 
                                     "8-cell", 
                                     "Morula", "Blastocyst (Mixed)", "Blastocyst (ICM)", "Heart", "Villus"),       
                           #split_by = "individual",
                           split_by = "Stage",
                           rm_low_quality = FALSE, # otherwise not enough samples
                           rm_cell_with_only_missing_values = TRUE,
                           cell_per_type = cell_per_type,
                           clustRow = FALSE,
                           h_width = 4,
                           h_height =5,
                           fs = 8)),  labels = c('B')))

figB<- cowplot::plot_grid(D,E, 
                   # rel_widths = c(2,1),
                   # scale = c(1, 0.8),
                   # label_size = 10, 
                   nrow = 1,
                   align = "h")

# pdf(file = paste0("../figs/Fig8C_", gene, "_Methylation_embryos.pdf"))#, height = 3, width = 3)
# figB
# dev.off()


figC <- cowplot::plot_grid(figA, figB,
                    rel_heights = c(1,1.6),
                   # scale = c(1, 0.8),
                   # label_size = 10, 
                   nrow = 2,
                   align = "h",
                   labels = my_gene)


pdf(file = paste0("../figs/Fig8C_", gene, "_Methylation_embryos.pdf"), height = 3.5, width = 4.5)
figC
dev.off()

figC
```










# Session info

```{r}
sessionInfo()
```


